---
title: "10X PBMC GEX Processing"
author: "Philipp SL Schäfer"
date: '`r format(Sys.time(), "%Y-%m-%d-%H-%M")`'
editor_options: 
  chunk_output_type: inline
bibliography: references.bib
csl: science.csl
link-citations: TRUE
---

# Status

Work in progress.

# Introduction

The goals are:

 * Visualize common quality control statistic.
 
 * Determine thresholds using the QC visualizations.
 
 * Subset the cell accordingly.

# Setup

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
Sys.setenv(RETICULATE_PYTHON =
             "/home/pschaefer/miniconda3/envs/r-reticulate/bin/python")
timestamp <- format(Sys.time(), "%Y-%m-%d-%H-%M")
s <- 4242
```

Loading Packages

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(magrittr)
  library(sparseMatrixStats)
  library(scDblFinder)
  library(scuttle)
  library(scran)
  library(scater)
  library(ggside)
})
```

# Import Data

```{r}
if (Sys.info()["nodename"] == "CrunchyPeanut") {
  input_dir <- "/mnt/sda/data/10X_pbmc_gex_10k/"
  output_dir <- "/home/philipp/Bioinfo_Playground/10x_pbmc_gex_processing_output/"
  if (!dir.exists(output_dir)) dir.create(output_dir)
} else {
  warning("No appropriate nodename, adjust paths manually")
  stop()
}
```

```{r}
list.files(input_dir)
```


# Cellranger QC Measures

We will also read in the summary.csv files generated by cellranger.

```{r}
read_csv(list.files(input_dir, full.names = TRUE)[
  str_detect(list.files(input_dir), "summary.csv$")], 
  show_col_types = FALSE) %>%
  mutate(across(everything(), as.character)) %>%
  pivot_longer(cols=everything(), names_to="Metric", values_to="Value")
```

Reading in the snRNA-seq data. We will read in the "raw_feature_bc_matrix.h5" matrices.

```{r}
rna_raw <- Seurat::Read10X_h5(
  list.files(input_dir, full.names = TRUE)[
  str_detect(list.files(input_dir), "raw_feature_bc_matrix.h5$")]
)
str(rna_raw)
```

# Filter Genes

First we will remove all genes that are not expressed in at least 3 cells.

```{r}
dim(rna_raw)
rna_raw <- rna_raw[sparseMatrixStats::rowSums2(rna_raw != 0) > 0, ]
dim(rna_raw)
```

# Empty Droplets

Before subsetting we will create a tibble containing the metadata which we will update throughout the analysis.

```{r}
coldata <- tibble::tibble(
  barcode = colnames(rna_raw),
  sample = "pbmc_1")
```

Plotting log10 of the cumulative UMI counts vs the rank of the cell or rather barcode ("kneeplot"). We need to find thresholds for each sample.

```{r}
purrr::map_dfr(unique(coldata$sample), function(smp) {
  tibble(sample = smp, 
         umi_per_cell = sparseMatrixStats::colSums2(rna_raw[, coldata$barcode[coldata$sample==smp]]))
  }) %>%
  arrange(sample, desc(umi_per_cell)) %>%
  group_by(sample) %>%
  mutate(idx = seq_along(sample)) %>%
  mutate(cum_umi_per_cell = cumsum(umi_per_cell)) %>%
  ggplot() +
  geom_line(aes(x=idx, y=cum_umi_per_cell)) +
  facet_wrap(~sample, scales = "fixed") +
  scale_y_log10() + scale_x_log10() +
  labs(x = "Index", y = "Cumulative UMI Count")
```

And we will be using the following thresholds (determined by visual inspection). Especially for "sm027_Tel_5c_20211123" there is a very obvious knee, for the other samples the thresholds are a little bit arbitrary (like everything in bioinformatics).

```{r}
thresholds <- c("pbmc_1" = 10.5e3)
thresholds
```

```{r}
p.list <- purrr::map(unique(coldata$sample), function(smp) {
  tibble::tibble(
    sample = smp,
    umi_per_cell = sparseMatrixStats::colSums2(rna_raw[, coldata$barcode[coldata$sample==smp]])) %>%
    arrange(sample, desc(umi_per_cell)) %>%
    group_by(sample) %>%
    mutate(idx = seq_along(sample)) %>%
    mutate(cum_umi_per_cell = cumsum(umi_per_cell)) %>%
    ggplot() +
    geom_line(aes(x=idx, y=cum_umi_per_cell)) +
    geom_vline(xintercept = thresholds[smp], color = "red") +
    scale_y_log10() + scale_x_log10() +
    labs(x = "Index", y = "Cumulative UMI Count", title = smp) +
    theme(plot.title = element_text(hjust = 0.5))
})
do.call(gridExtra::grid.arrange, p.list)
```

Lastly we subset the count matrices based on these results.

```{r}
cell_qc <- purrr::map_dfr(unique(coldata$sample), function(smp) {
  tibble(sample = smp, 
         sum_umis = sparseMatrixStats::colSums2(rna_raw[, coldata$barcode[coldata$sample==smp]]),
         detected_genes = sparseMatrixStats::colSums2(rna_raw[, coldata$barcode[coldata$sample==smp]] != 0),
         barcode = coldata$barcode[coldata$sample==smp], 
         threshold = thresholds[smp]) }) %>%
  arrange(sample, desc(sum_umis)) %>%
  group_by(sample) %>%
  mutate(rank = seq_along(sample)) %>%
  mutate(cum_umi_per_cell = cumsum(sum_umis)) %>%
  mutate(non_empty = ifelse(rank <= threshold, TRUE, FALSE))

coldata <- coldata %>%
  dplyr::left_join(cell_qc, by=c("sample", "barcode")) %>%
  dplyr::select(sample, barcode, sum_umis, detected_genes, non_empty)

dim(rna_raw)
rna <- rna_raw[,coldata %>% dplyr::filter(non_empty) %>% dplyr::pull(barcode)]
coldata <- coldata %>% dplyr::filter(non_empty)
dim(rna)
```

Check again. And I think it looks ok now.

```{r}
purrr::map_dfr(unique(coldata$sample), function(smp) {
  tibble(sample = smp, 
         umi_per_cell = sparseMatrixStats::colSums2(rna[, coldata$barcode[coldata$sample==smp]]))
  }) %>%
  arrange(sample, desc(umi_per_cell)) %>%
  group_by(sample) %>%
  mutate(idx = seq_along(sample)) %>%
  mutate(cum_umi_per_cell = cumsum(umi_per_cell)) %>%
  ggplot() +
  geom_line(aes(x=idx, y=cum_umi_per_cell)) +
  facet_wrap(~sample, scales = "fixed") +
  scale_y_log10() + scale_x_log10() +
  labs(x = "Index", y = "Cumulative UMI Count")
```

# RNA Doublet Scores - `scDblFinder`

We will use `scDblFinder` as this package it implemented in R and seems to be better than `scrublet` (according to their vignette at least, but I guess their comparison might be biased!). Usually these tools simulate doublets and check how similar your actual cells are to these simulated doublets.

See the description/vignette of the tool here: 

1. http://www.bioconductor.org/packages/release/bioc/vignettes/scDblFinder/inst/doc/computeDoubletDensity.html

2. http://www.bioconductor.org/packages/release/bioc/vignettes/scDblFinder/inst/doc/scDblFinder.html

To use the tools we will temporarily create a single cell experiment object.

```{r fig.width=10, fig.height=4, message=FALSE}
dbr <- 0.076
print(paste0("Using an expected doublet rate of: ", dbr))

sce <- SingleCellExperiment(list(counts = rna))
sce <- logNormCounts(sce)
dec <- modelGeneVar(sce)
hvgs <- getTopHVGs(dec, n=2000)
set.seed(s)
sce <- runPCA(sce, ncomponents=30, subset_row=hvgs)
sce <- runUMAP(sce, dimred="PCA")
#scores <- computeDoubletDensity(sce, subset.row=hvgs)
sce <- scDblFinder(sce, dbr=dbr, iter=1)
scores <- sce$scDblFinder.score
classification <- sce$scDblFinder.class
barcode <- rownames(sce@colData)

# prepare plots
p1 <- tibble::tibble(cell = names(scores), score = scores, 
               UMAP1 = reducedDim(sce, "UMAP")[,1],
               UMAP2 = reducedDim(sce, "UMAP")[,2]) %>%
  ggplot() +
  geom_point(aes(x=UMAP1, y=UMAP2, color=score), size=0.2) +
  scale_color_viridis_c()
p2 <- ggplot() +
  geom_histogram(aes(scores), bins = 100, fill="grey", color="black") +
  labs(y = "Count", x = "Doublet Score")
  
# use cowplot for plotting the grid
cowplot::plot_grid(p1, p2)
```


```{r fig.width=10, fig.height=4, message=FALSE}
doublet_scores <- tibble::tibble(barcode = barcode,
                                 score = scores,
                                 classification = classification)
coldata <- coldata %>%
  left_join(doublet_scores, by="barcode")
```

Check how many predicted doublets per sample using our threshold set above. Let's keep the doublet scores and we will not remove/mask any cells, but rather check later on whether doublets are particularly enriched in any cluster we get.

```{r}
coldata %>%
  dplyr::count(sample, classification)
```

# QC Statistics

Let's look at some common QC statistics:

1. Total number of UMIs per cell
2. Total number of detected genes per cell
3. Fraction of mitochondrial genes.

However, using mitochondrial genes probably makes no sense looking at single-nuc data. See this paper for a short comparison between single-cell and single-nuc RNA-sequencing. [@10.1042/etls20210074] Furthermore, we do not know the names of the mitochondrial gene here, most likely they are gene starting with "mt-/_". 

```{r}
mito_genes <- 
  rownames(rna)[grep("^MT-", rownames(rna))]
mito_genes
```

Computing the QC stats.

Looking at the bottom/left tail of the distributions one might argue that the cutoff with regards to the number of UMIs is a little bit too strict for sample "sm023".

```{r}
qc_stats_before <- tibble::tibble(
  sample = "pcmbc_1",
  barcode = colnames(rna),
  umis_per_cell = Matrix::colSums(rna),
  genes_per_cell = Matrix::colSums(rna != 0),
  mt_fraction = Matrix::colSums(rna[mito_genes, ]) /  Matrix::colSums(rna) * 100
)

qc_stats_before %>%
  pivot_longer(cols = !c(sample, barcode), names_to = "stat") %>%
  mutate(stat = ifelse(stat=="mt_fraction", "mt_fraction (%)", stat)) %>%
  ggplot() +
  geom_violin(aes(x=sample, y=value, fill=sample)) +
  facet_wrap(~ stat, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

Generally speaking the initial QC should not be too stringent.

Removing cells with:

1. More than 9000 genes per cell
2. Mitochondrial fraction (%) above 15
3. More than 20000 UMIs per cell

Let's add the QC thresholds to our coldata.

```{r}
coldata <- coldata %>%
  mutate(sum_umis_in_bound = (sum_umis <= 20000),
         detected_genes_in_bound = (detected_genes <= 9000),
         mt_fraction = Matrix::colSums(rna[mito_genes, ]) /  
           Matrix::colSums(rna) * 100,
         mt_fraction_in_bound = mt_fraction <= 15) %>%
  rowwise() %>%
  mutate(qc_pass = all(non_empty, sum_umis_in_bound, 
                       detected_genes_in_bound, mt_fraction_in_bound)) %>%
  ungroup()

coldata %>%
  dplyr::filter(qc_pass) %>%
  glimpse()
```

```{r}
qc_stats_after <- tibble::tibble(
  sample = "pcmbc_1",
  barcode = colnames(rna[,coldata$barcode[coldata$qc_pass]]),
  umis_per_cell = Matrix::colSums(rna[,coldata$barcode[coldata$qc_pass]]),
  genes_per_cell = Matrix::colSums(rna[,coldata$barcode[coldata$qc_pass]] != 0),
  mt_fraction = Matrix::colSums(rna[,coldata$barcode[coldata$qc_pass]][mito_genes, ]) /  Matrix::colSums(rna[,coldata$barcode[coldata$qc_pass]]) * 100
)

qc_stats_after %>%
  pivot_longer(cols = !c(sample, barcode), names_to = "stat") %>%
  mutate(stat = ifelse(stat=="mt_fraction", "mt_fraction (%)", stat)) %>%
  ggplot() +
  geom_violin(aes(x=sample, y=value, fill=sample)) +
  facet_wrap(~ stat, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

# Saving

```{r message=FALSE}
saveRDS(rna, 
        paste0(output_dir, timestamp, "10x_pbmc_processed_rna_counts.RDS"))
saveRDS(coldata, 
        paste0(output_dir, timestamp, "10x_pbmc_processed_coldata.RDS"))

save.image(file = paste0(output_dir, "Latest_10x_pbmc_processed.RData"))
```

# References
